---
title: "Draws with a dynamic tolerance"
author: "Petr Fedorov"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Draws with a dynamic tolerance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
  - id: gerlach2019
    title: Dissection of Bitcoin’s multiscale bubble history from January 2012 to February 2018
    author: 
      - family: Gerlach
        given: J.C.
      - family: Demos
        given: G.
      - family: Sornette
        given: D.
    issued:
      year: 2019
      month: 7
    container-title: Royal Society Open Science
    type: article-journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tibble)
library(ggplot2)
library(latex2exp)

```

# $\epsilon$-draws: draws with a static tolerance

Nowadays $\epsilon$-drawdowns and $\epsilon$-drawups are being acitvely used in the scientific literature and one of the most recent examples is the article (@gerlach2019). The description of $\epsilon$-drawdowns /$\epsilon$-drawups methodology that follows is based on the article. 

Consider we are given a price series $P[t_i], \quad i=0,1, \ldots$ of some kind. For example, it could be the consecutive "best ask" prices of BTCUSD pair at some exchange.  $t_i$ deisgnates the moment when the price has changed (i.e. a new best order has arrived or the previous best order has been cancelled).
We can calculate the discrete log-returns 
$$
\begin{equation} 
r_i = \ln P[t_i] - \ln P[t_{i-1}], \quad i=1,2,\ldots (#eq:a1)
\end{equation}
$$

The first time $t_{i_0}$ is defined as the beginning of a drawup (drawdown) if $r_1>0$($r_1 < 0$).
Then, for each subsequent $t_{i_0+i} > t_{i_0}, \quad i=1,2,\ldots$, we calculate the cummulative return up to $t_{i_i}$ as 
$$
\begin{equation}
p_{i_0,i} = \sum_{k=1}^i r_{i_k} = \ln P[t_{i_i}] - \ln P[t_{i_0}] (#eq:a2)
\end{equation}
$$

At each time $t_{i_i}, \quad i=1,2,\ldots$ we need to check whether the current drawup (drawdown) phase is still active. We test this by calculating the largest deviation $\delta_{i_0,i}$ of the price trajectory from a previous maximum (minimum) 
$$
\begin{equation}
\delta_{i_0,i} = \begin{cases} \max_{i_0 \leq k \leq i} \{p_{i_0,k}\} - p_{i_0,i} & \text{for drawups} \\  p_{i_0, i}  - \min_{i_0 \leq k \leq i} \{p_{i_k,i}\}& \text{for drawdowns} \end{cases} (#eq:a3)
\end{equation}
$$
The procedure is stopped at time $i$ when the deviation exceeds a predefined tolerance $\epsilon$ 
$$
\begin{equation}
\delta_{i_0,i} > \epsilon (#eq:a4)
\end{equation}
$$ The stopping tolerance quantifies how much the price is allowed to move in the direction opposite to the
drawup/drawdown trend.
When the procedure has been stopped, the end of the current drawup (drawdown) phase is determined as the time of the highest (lowest) price seen in the tested interval: 
$$
\begin{equation}
i_1 = \begin{cases} \arg \max_{i_0 \leq k \leq i} \{p_{i_0,k}\} & \text{for drawups} \\ \arg \min_{i_0 \leq k \leq i} \{p_{i_0,k}\}& \text{for drawdowns} \end{cases} (#eq:a5)
\end{equation}
$$
The $\epsilon$-drawup/drawdown procedure \@ref(eq:a5) is restarted at time $i_1$. The start of the next drawup/drawdown period will then be located at this time. By construction of $\delta$ and the stopping condition a drawup (respectively, drawdown) is always followed by a drawdown (respectively,drawup). The procedure is repeated until the full length of the analysed time series is represented as a sequence of drawups and drawdowns. The process is illustrated on figure \@ref(fig:epsilon-draws-example).


```{r epsilon-draws-example, echo=FALSE, fig.align="center", fig.cap="Here the $\\epsilon$-draw methodology is illustrated for the case when the drawup stopped at time $t_{i_0+3}$ because   $\\delta_{i_0, 5}$ exceeded $\\epsilon$ and the drawdown then started. Note that the drawup didn't stop at time $t_{i_0+1}$ since the tolerance $\\epsilon$ was not exceeded."}
eps.draws.data <- tibble(P=c(10, 13,12, 18, 16, 14), T= c(1,4, 5,6,7.5, 8), k=seq_along(T))
eps.draws.horiz <- with(eps.draws.data, tibble(x=c(T[1], T[2], T[4],  T[6]), y=c(P[1], P[2], P[4], P[6]), xend=c(T[6]+2, T[2]+0.5,T[6]+2,  T[6]+2), yend=y))

eps.draws.vert <- with(eps.draws.horiz, tibble(x=c(xend[2],xend[4], xend[4] ), y=c(!!y[1], !!y[4], !!y[1]), xend=c(xend[2], xend[4],xend[4]),yend=c(!!y[2], !!y[3], !!y[4]), t=c(TeX("$r_{i_0+1}$", output="character"), TeX("$\\delta_{i_0,5} > \\epsilon$",output="character"), TeX("$p_{i_0,5}$",output="character"))))
eps.draws <- with(eps.draws.data, tibble(x=c(T[1], T[4]), y=c(P[1], P[4]), xend=c(T[4],T[6]), yend=c(P[4], P[6]), c=c("green", "red")))

ggplot(eps.draws.data, aes(x=T, y=P)) + 
  geom_point() + geom_line() + 
  geom_segment(data=eps.draws.horiz, aes(x=x, y=y, xend=xend, yend=yend), size=0.2) +
  geom_segment(data=eps.draws.vert, aes(x=x, y=y, xend=xend, yend=yend), arrow = arrow(ends="both",angle = 10, length = unit(0.5, "lines")), size=0.2) +
  geom_text(data=eps.draws.vert, aes(x=x+0.1, y=(y+yend)/2, label=t, hjust=0), parse=TRUE) +
  scale_x_continuous(TeX("$t_i$"),breaks=eps.draws.data$T, labels = c(TeX("$t_{i_0}$"),TeX("$t_{i_0+1}$"),TeX("$t_{i_0+2}$"),TeX("$t_{i_0+3} = t_{i_1}$"),TeX("$t_{i_0+4}$"),TeX("$t_{i_0+5}$") )) +
  geom_segment(data=eps.draws, aes(x=x, y=y, xend=xend, yend=yend, colour=c), arrow = arrow(angle=10,length = unit(1, "lines"))) +
  expand_limits(x=11) + 
  scale_y_continuous(TeX("$\\ln P\\[t_i\\]$"), breaks=eps.draws.data$P, labels = c(TeX("$\\ln P\\[ t_{i_0}\\]$"),TeX("$\\ln P\\[t_{i_0+1}\\]$"),TeX("$\\ln P\\[t_{i_0+2}\\]$"),TeX("$\\ln P\\[t_{i_0+3}\\]$"),TeX("$\\ln P\\t_{i_0+4}\\]$"),TeX("$\\ln P\\[t_{i_0+5}\\]$") )) + 
  scale_color_manual("Draw direction", values=c("red"="red", "green"="green"), labels=c("up", "down")) +
  theme(legend.position = "top")

  

```


# The drawback of $\epsilon$-draws methodology

As the authors of (@gerlach2019) point out  it is difficult to single out  an appropriate $\epsilon$ value.  

> Selections resulting in large values of the tolerance $\epsilon$ will tend to yield a coarse long-term
sequence while small $\epsilon$-values result in more frequent interruption of a drawup or drawdown,
yielding finer sequences.

Thus in both cases it is not possible to get an intuitevely expected **mix** of large and small draws. Instead either all your draws are large (i.e. when $\epsilon$ is large) or small (when it is small). This is the drawback of $\epsilon$-draws methodology to be addressed by a dynamic $\epsilon$-draws.


# A dynamic $\epsilon$-draws methodology

In order to address the issue mentioned above, we need to adjust $\epsilon$ in formula \@ref(eq:a4) dynamically 
$$
\begin{equation}
\epsilon = \begin{cases} \max_{i_0 \leq k \leq i} \{p_{i_0,k}\}\gamma_{i_0,i}  & \text{for drawups} \\  - \min_{i_0 \leq k \leq i} \{p_{i_k,i}\}\gamma_{i_0,i} & \text{for drawdowns} \end{cases} (#eq:d1) 
\end{equation}
$$
First, note that $\epsilon$ in \@ref(eq:d1) is now directly proportional to the size of the latest draw, so the larger the draw the bigger must be the draw in the opposite direction to stop it.  

Then $\gamma_{i_0,i}$ is calculates using the following formula
$$
\gamma_{i_0,i} = \frac{\gamma_0/100}{1 + \theta(t_{i_0+i} - t_{i_0})} 
$$

where $\gamma_0$ is the minimal draw percentage coefficient and $\theta$ is a time decay coefficient. Both $\gamma_0$ and $\theta$ are constants.

If $\theta = 0$ then $\gamma_{i_0,i} = \gamma_0$. The draw will be stopped by the next draw which is no less in size than $\gamma_0$ percents of the current draw. 

If $\theta > 0$ then $\gamma_{i_0,i}$ will decay with time giving a very large draw a better chance to stop. 


# References
