---
title: An analysis of the correlations between LTCUSD price movements at Bitstamp and Bitfinex 
author: "Petr Fedorov"
date: 'September 22, 2019'
output: html_document
params:
  start_time: 2019-09-01 00:00:00+03
  end_time: 2019-09-22 00:00:00+03
  exchange1: bitfinex
  exchange2: bitstamp
  pair: LTCUSD
  interval1: 2
  interval2: 10
  interval3: 30
  depth_start_time: 2019-09-21 00:20:00+03
  depth_end_time: 2019-09-21 00:30:00+03
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(latex2exp)
library(knitr)
library(gsubfn)

options(digits.secs = 4, scipen=999)


default_source_hook <- knit_hooks$get('source')

knit_hooks$set(source = function(x, options) {
  x <- gsubfn(x = x, pattern = "params\\$`?([\\w_]+)`?", function(y) {
    y <- get(y, params)
    ifelse(is.character(y), paste0("'", y, "'"), y)
  })
  default_source_hook(x, options)
})


```


```{r include=FALSE}
config <- config::get()
con <- obadiah::connect(user=config$user,dbname=config$dbname, host=config$host,
                      port=config$port, sslrootcert=config$sslrootcert,
                      sslcert=config$sslcert,sslkey=config$sslkey)


```

# Introduction

On the figure below the white line shows how the mid-price of LTCUSD pair changes at Bitfinex and Bitstamp between `r params$depth_start_time` and `r params$depth_end_time`. The changes look similar but they are not completely the same. One could ask: is it possible to use the mid-price change that *has just happened* at one of the exchanges to predict the change that *will hapen* at the other? Let's try to answer this question.

```{r fig.align="center", echo=FALSE, cache=TRUE}
gridExtra::grid.arrange(
  obAnalytics::plotPriceLevels(
    obadiah::depth(con, 
                   params$depth_start_time,
                   params$depth_end_time,
                   params$exchange1,
                   params$pair,
                   tz='Europe/Moscow'),
    obadiah::spread(con, 
                    params$depth_start_time,
                    params$depth_end_time,
                    params$exchange1,
                    params$pair,
                    tz='Europe/Moscow')
  ) + ggtitle(paste0(params$exchange1, ',', params$pair)),
  obAnalytics::plotPriceLevels(
    obadiah::depth(con, 
                   params$depth_start_time,
                   params$depth_end_time,
                   params$exchange2,
                   params$pair,
                   tz='Europe/Moscow'),
    obadiah::spread(con, 
                    params$depth_start_time,
                    params$depth_end_time,
                    params$exchange2,
                    params$pair,
                    tz='Europe/Moscow')
  )+ ggtitle(paste0(params$exchange2, ',', params$pair))
)  
```


# Method

We will sample LTCUSD mid-price at regular intervals (`r params$interval1`, `r params$interval2` and `r params$interval3` seconds) at Bitstamp and Bitfinex from `r params$start_time` till `r params$end_time`. Then we'll calculate mid-price change during each interval. Finally we'll estimate Pearson correlation between mid-price change at Bitstamp and at Bitfinex using intervals of the same length but shifted relative to each other in time. We'll also look at the size of mid-price changes.

We start from creating a tibble whith definitions of two instruments we will compare. It will be used to call functions from obadiah package that are vectorized over instruments:


```{r}
instruments <- tribble(~exchange, ~pair,
                       params$exchange1, params$pair,
                       params$exchange2, params$pair)
```


The first such function allows us to check whether we have required data in the database:

```{r fig.align="center"}
obadiah::plotDataAvailability( 
  with(instruments, obadiah::intervals(con, params$start_time, params$end_time, exchange, pair, tz='Europe/Moscow')) 
)
```

Almost everything is green, so we may start the analysis.

`obadiah::spread()` will load from the database the spread the the end of the interval specified `by.interval` parameter. The function is vectorized over `exchange` and `pair` parameters so we utilize our `instruments` here.  We will analyse several intervals held in `intervals` vector:



```{r, cache=TRUE}
intervals <- c( params$interval1, params$interval2, params$interval3)  # seconds 

mid.price.changes <- purrr::map_dfr(intervals, 
                                 function(d) {
                                   with(instruments,
                                        obadiah::spread(con,
                                                        params$start_time,
                                                        params$end_time,
                                                        exchange=exchange,
                                                        pair=pair,
                                                        frequency=d,
                                                        tz='Europe/Moscow')) %>% 
                                     group_by(exchange, pair) %>%
                                     mutate(mid.price.change=log10(best.bid.price+best.ask.price) -
                                                             log10(lag(best.bid.price)+lag(best.ask.price)),
                                            instrument=paste0(pair,'_',exchange),
                                            duration=!! enquo(d)) %>%
                                     filter(!is.na(mid.price.change))  %>% 
                                     ungroup() %>%  
                                     select(duration, timestamp, mid.price.change, instrument) %>%
                                     tidyr::spread(instrument, mid.price.change)  # Note that it is tidyr::spread() - spreads key-value pair accross multiple columns
                                   }
                                 )
kable(tail(mid.price.changes))

```

The number of non-NA observations is significant for all intervals:

```{r}

kable(mid.price.changes %>%
  group_by(duration) %>%
  summarize(non_NA_bitstamp=sum(!is.na(LTCUSD_bitstamp)), non_NA_bitfinex=sum(!is.na(LTCUSD_bitfinex))), col.names=c("Duration of interval", "Bitstamp", "Bitfinex"), align=c("c", 'c', 'c'), caption="Number of non-NA mid-price changes")
  

```


We will calculate correlation between `LTCUSD_bitstamp` and `lead(LTCUSD_bitfinex,n)` or `lag(LTCUSD_bitfinex,-n)` where `n` is the number of periods ahead or behind Bitstamp period respectively.

```{r}
correlations <- mid.price.changes %>%
  group_by(duration) %>%
  group_modify( function(sc, d) {
    purrr::map_dfr(-5:5, function(n) {
      if(n >= 0)
        with(sc, tibble(n=n, correlation=cor(LTCUSD_bitstamp, lead(LTCUSD_bitfinex,n), use="complete.obs")))
      else      
        with(sc, tibble(n=n, correlation=cor(LTCUSD_bitstamp, lag(LTCUSD_bitfinex,-n), use="complete.obs")))
      })
    }
    )
```


# Results

The estimated correlations are visualized below:

```{r fig.align="center"}
ggplot(correlations %>% 
         filter(abs(n*duration) <= 30 & abs(n) <= 3),
       aes(x=n*duration, 
           y=correlation, 
           group=duration, 
           colour=as.factor(duration))
       ) + 
  geom_line() + 
  geom_point() + 
  scale_y_continuous("Correlation between mid-price changes", 
                     labels = scales::percent,
                     breaks=seq(0,1,0.1), 
                     minor_breaks=seq(-1,1,0.01)) + 
  scale_x_continuous("Time relative to the Bitstamp's interval end, seconds", 
                     breaks=c(seq(-40,40,10), seq(-10,10,2)),
                     minor_breaks =seq(-40,40,1) ) + 
  geom_text(aes(label=paste0(duration, ':', n)), 
            vjust=0,
            hjust=0, 
            nudge_x=1, 
            nudge_y=-0.01, 
            size=3) + 
  scale_color_discrete("Duration of intervals, seconds") + 
  theme(legend.position = "bottom")

```

We see that:

1. The strongest correlation is between mid-price changes happened during simultaneous 30-seconds intervals at Bitstamp and Bitfinex (point 30:0).

2. The correlation is getting lower when the interval diminishes and becomes weak for 2-seconds intervals (point 2:0). 

3. There is no correlation between the mid-price change during 30-seconds interval at Bitstamp  and *next* 30-seconds interval at Bitfinex (point 30:1). 

4. There is a weak correlation between the mid-price change during 10-seconds and especially 2-seconds interval at Bitstamp and *next* and *previous* intervals of the same duration at Bitfinex (points 10:1, 10:-1, 2:1, 2:-1).



The scatter plots below show the relative size of mid-price changes during 2- and 30- seconds intervals. Surprisingly, they are almost the same.


```{r fig.align="center"}
ggplot(mid.price.changes %>%
         filter(duration %in% c(params$interval1, params$interval3)) %>%
         mutate(LTCUSD_bitfinex=lead(LTCUSD_bitfinex, n=1)) %>%
         filter(!is.na(LTCUSD_bitstamp) & !is.na(LTCUSD_bitfinex)),
       aes(x=LTCUSD_bitstamp, y=LTCUSD_bitfinex)) + 
  geom_point(shape='.') + 
  scale_x_continuous(labels=scales::percent) +
  scale_y_continuous(labels=scales::percent) +
  coord_fixed(xlim = c(-0.003, 0.003), ylim=c(-0.003, 0.003)) +
  geom_smooth(method="lm") + 
  labs(x="Mid-price change that has happened at Bitstamp", y="Mid-price change that will happen at Bitfinex") +
  facet_grid(cols=vars(duration))
```

# Discussion

At the time of writing the highest taker's fee at Bitfinex was 0.2% while at Bitstamp it was 0.5%, i.e. comparable or even greater than the maximum mid-price change during 2-seconds interval. This makes profitable arbitrage strategy utilizing the correlation discovered unlikely.
