---
title: "A theory of power-law distributions in financial market fluctuations"
#author: "Petr Fedorov"
#date: "`r Sys.Date()`"
output: html_document
params:
  exchange: Bitstamp
  pair: BTCUSD
  start_time: 2019-02-01 00:00:00+03
  end_time: 2019-03-21 00:00:00+03
  period: 15 min
references:
- id: Gabaix2003
  title: A theory of power-law distributions in financial market fluctuations
  author:
  - family: Gabaix
    given: Xavier
  - family: Gopikrishnan
    given: Parameswaran
  - family: Plerou
    given: Vasiliki
  - family: Stanley
    given: H. Eugene
  container-title: Nature
  volume: 423
  URL: 'https://www.nature.com/articles/nature01624'
  DOI: 10.1038/nature01624
  issue: 6937
  publisher: Nature Research
  page: 267-270
  type: article-journal
  issued:
    year: 2003
    month: 5
csl: institute-of-mathematical-statistics.csl      
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(latex2exp)
library(knitr)
library(gsubfn)

config <- config::get()
con <- DBI::dbConnect(RPostgres::Postgres(), user=config$user,dbname=config$dbname, host=config$host,
                      port=config$port, sslmode="require", sslrootcert=config$sslrootcert,
                      sslcert=config$sslcert,sslkey=config$sslkey,bigint="numeric")

cache <- new.env(parent=emptyenv())
options(digits.secs = 4, scipen=999)


default_source_hook <- knit_hooks$get('source')

knit_hooks$set(source = function(x, options) {
  x <- gsubfn(x = x, pattern = "params\\$`?([\\w_]+)`?", function(y) {
    y <- get(y, params)
    ifelse(is.character(y), paste0("'", y, "'"), y)
  })
  default_source_hook(x, options)
})


```

According to [@Gabaix2003]:

> Power laws appear to describe histograms of relevant financial fluctuations, such as fluctuations in stock price, trading volume and the number of trades. **Surprisingly, the exponents that characterize these power laws are similar for different types and sizes of markets, for different market trends and even for different countries—suggesting that a generic theoretical basis may underlie these phenomena.**

Let's check whether these exponents are similar for cryptocurrencies too. Spoiler: they are not.




## Trades

In [@Gabaix2003] we read:

> Empirical studies ... show that the distribution of trading volume $V_t$ obeys a ... power law: $$ P(V_t > x) \sim x^{-\zeta_V}$$ with $\zeta_V \approx 1.5$, while the number of trades $N_t$ obeys: $$ P(N_t > x) \sim x^{-\zeta_N}$$ with $\zeta_N \approx 3.4$


First let's load all trades for **`r params$pair`** pair traded at **`r params$exchange`** from `r params$start_time` till `r params$end_time`:


```{r, cache=TRUE}
trades <- obadiah::trades(con, 
                          params$start_time, 
                          params$end_time, 
                          params$exchange, 
                          params$pair, 
                          cache=cache,  tz='Europe/Moscow') %>%
  select(-maker.event.id, -taker.event.id)
kable(head(trades))
```

We need only **real** trades, i.e. those having `exchange.trade.id` set and only `timestamp` and `volume` columns:

```{r}
trades <- trades %>% 
  filter(!is.na(exchange.trade.id)) %>% 
  select(timestamp, volume, price, direction, exchange.trade.id)
kable(head(trades))
```

There are `r scales::number_format(big.mark=',')(nrow(trades))` **real** trades in total. Let's calculate a trading volume and number of trades per `r params$period` period:


```{r}
by_period <- left_join(tibble(period_end=seq(ceiling_date(ymd_hms(params$start_time), params$period),
                                                 ceiling_date(ymd_hms(params$end_time), params$period),
                                                 by=params$period)),
                           trades %>% 
                             mutate(period_end=ceiling_date(timestamp, params$period))
                           , by="period_end") %>% 
  mutate(volume=replace_na(volume, 0)) %>%
  group_by(period_end) %>%
  summarize(volume=sum(volume), number.of.trades=sum(!is.na(timestamp)))

knitr::kable(head(by_period))

```

We can now caclulate [Complementary cumulative distribution function (tail distribution)](https://en.wikipedia.org/wiki/Cumulative_distribution_function#Complementary_cumulative_distribution_function_(tail_distribution)) for a trading volume per period $P(V_t > x)$:


```{r}
V <- by_period %>% 
  arrange(-volume) %>%
  mutate(r=row_number()) %>%
  mutate(prob=r/max(r), sigma=volume/sd(volume)) %>%
  filter(sigma >= 0.1)

s_i <- coef(lm(log10(prob) ~ log10(sigma), V %>% filter(sigma >=2)))
zeta_V <- -s_i[2]
```

$x$ is measured in units of sample standard deviation of `volume`: $\sigma_V =$`sd(volume)` $\approx$ `r round(sd(by_period$volume),2)`.  



```{r, fig.width=7}
ggplot(V ,  aes(sigma, prob)) + 
  geom_point(size=0.1) + 
  scale_y_log10(TeX('$P(V_t > x)$'),
                breaks=c(0.001, 0.02, 0.1, 1),
                labels=scales::percent) +
  scale_x_log10(TeX('$\\frac{x}{\\sigma_N$}'),
                breaks=c(0.1, 1,2,10, 100,40),
                limits=c(0.1, 40),
                labels=scales::number_format(0.1)) +
  geom_abline(slope=-zeta_V, intercept=s_i[1], colour="blue") + 
  annotate("text",
           x=10,
           y=0.01, 
           label=TeX(paste0('$\\zeta_V=',round(zeta_V,2),'$'), output = "character"),
           parse=TRUE)

```


We see that $\zeta_V=`r round(zeta_V,2)` \neq 1.5$. 

For a number of trades per period $P(N_t > x)$ we have:


```{r}
N <- by_period %>% 
  arrange(-number.of.trades) %>%
  mutate(r=row_number()) %>%
  mutate(prob=r/max(r), sigma=number.of.trades/sd(number.of.trades), category='N') %>%
  filter(sigma > 0.1)

s_i <- coef(lm(log10(prob) ~ log10(sigma), N %>% filter(sigma >=2)))
zeta_N <- -s_i[2]

```

Again, $x$ is measured in units of sample standard deviation of `number.of.trades`: $\sigma_N =$`sd(number.of.trades)` $\approx$ `r round(sd(by_period$number.of.trades),2)`.  



```{r, fig.width=7}
ggplot(N ,  aes(sigma, prob)) + geom_point(size=0.1) + 
  scale_y_log10(TeX('$P(N_t > x)$'),
                breaks=c(0.001, 0.01, 0.1, 1),
                labels=scales::percent) + 
  scale_x_log10(TeX('$\\frac{x}{\\sigma_N$}'),
                breaks=c(0.1, 1, 2, 10, 40),
                limits=c(0.1, 40),
                labels=scales::number_format(0.01)) +
  geom_abline(slope=-zeta_V, intercept=s_i[1], colour="blue") +
  annotate("text",
           x=10, 
           y=0.01, 
           label=TeX(paste0('$\\zeta_N=',round(zeta_N,2),'$'), output = "character"),
           parse=TRUE)

```


We see that $\zeta_N=`r round(zeta_N,2)` \neq 3.4$. 

## Returns


In [@Gabaix2003] we read:

> Define $p_t$ as the price of a given stock and the stock price ‘return’ $r_t$ as the change of the logarithm of stock price in a given time interval $\Delta t$, $r_t \equiv \ln p_t - \ln p_{t -\Delta t}$. The probability that a return has an absolute value larger than $x$ is found empirically to be: $$ P(| r_t | > x) \sim x^{-\zeta_r} \tag{1}$$ with $\zeta_r \approx 3$. The ‘inverse cubic law’ of equation (1) is rather ‘universal’, holding over as many as 80 standard deviations for some stock markets, with $\Delta t$ ranging from one minute to one month, across different sizes of stocks, different time periods, and also for different stock market indices.


Let's load all level 1 change records for **`r params$pair`** pair traded at **`r params$exchange`** from `r params$start_time` till `r params$end_time`:


```{r, cache=TRUE}
spread <- obadiah::spread(con, 
                          params$start_time, 
                          params$end_time, 
                          params$exchange, 
                          params$pair, 
                          cache=cache,  tz='Europe/Moscow')
kable(head(spread))
```

We will use a *mid-price* as a proxy for $p_t$:

```{r}
spread <- spread %>%
  mutate(mid.price=(best.ask.price+best.bid.price)/2) %>%
  select(timestamp, mid.price)
kable(head(spread))  
```


```{r}
by_period <- left_join(tibble(period_end=seq(ceiling_date(ymd_hms(params$start_time), params$period),
                                                 ceiling_date(ymd_hms(params$end_time), params$period),
                                                 by=params$period)),
                           spread %>% 
                             mutate(period_end=ceiling_date(timestamp, params$period))
                           , by="period_end") %>%
  group_by(period_end) %>%
  summarize(price=last(mid.price, order_by=timestamp))  %>%
  mutate(abs_r=abs(log(price) - log(lag(price))))  %>% 
  filter(!is.na(abs_r))

kable(head(by_period))
```



```{r}
R <- by_period %>%
  arrange(-abs_r) %>%
  mutate(r=row_number()) %>%
  mutate(prob=r/max(r), sigma=abs_r/sd(abs_r)) %>%
  filter(sigma >= 0.1)

s_i <- coef(lm(log10(prob) ~ log10(sigma), R %>% filter(sigma >=2)))
zeta_r <- -s_i[2]


```

$x$ is measured in units of sample standard deviation of `abs_r`: $\sigma_r =$`sd(abs_r)` $\approx$ `r round(sd(by_period$abs_r),4)`.  


```{r, fig.width=7}
ggplot(R ,  aes(sigma, prob)) + 
  geom_point(size=0.1) + 
  scale_y_log10(TeX('$P(|r_t| > x)$'),
                breaks=c(0.001, 0.02, 0.1, 1),
                labels=scales::percent) +
  scale_x_log10(TeX('$\\frac{x}{\\sigma_r$}'),
                breaks=c(0.1, 1,2,10, 100,40),
                limits=c(0.1, 40),
                labels=scales::number_format(0.1)) +
  geom_abline(slope=-zeta_r, intercept=s_i[1], colour="blue") + 
  annotate("text",
           x=10,
           y=0.01, 
           label=TeX(paste0('$\\zeta_r=',round(zeta_r,2),'$'), output = "character"),
           parse=TRUE)

```

Again we see that $\zeta_r=`r round(zeta_r,2)` \neq 3$. 

## References

